% =========================================================================
% T20d: Corrected finite-size-scaling status for 3D Z_2 lattice gauge theory
%
% This file preserves the post-May simulation record while withdrawing the
% unsupported first-order interpretation. It is not included in timesarrow.tex.
% =========================================================================

\section{Finite-Size Scaling Status of the 3D \texorpdfstring{$\mathbb{Z}_2$}{Z2} Transition}
\label{sec:t20d-fss}

The three-dimensional pure $\mathbb{Z}_2$ lattice gauge theory is dual to the
three-dimensional Ising model. Its confinement--deconfinement transition is
therefore continuous and belongs to the 3D Ising universality class. The
post-May simulations locate rapid finite-size variation near
$\beta\simeq0.76$, consistent with the established critical region, but the
present measurements do not constitute an independent precision determination
of the critical coupling or critical exponents.

\subsection{What Was Measured}

The retained data contain the mean elementary plaquette, its fluctuations, a
plaquette Binder ratio, and selected Wilson loops for lattices up to $L=32$.
The elementary plaquette is a local, energy-like observable. It is not a
Wilson-loop order parameter: confinement is diagnosed from the size dependence
of extended Wilson loops, for example through an area-law or perimeter-law fit.

For the plaquette variable $P$, the stored fluctuation observables were defined
as
\begin{align}
 \chi_P &= L^3\bigl(\langle P^2\rangle-\langle P\rangle^2\bigr),\\
 U_P &= 1-\frac{\langle P^4\rangle}{3\langle P^2\rangle^2}.
\end{align}
Because $\langle P\rangle$ is nonzero, a narrow plaquette distribution
generically gives $U_P\to2/3$. This limit is not a universal Binder crossing
and does not distinguish a continuous transition from a first-order one.

\subsection{Corrected Interpretation}

The earlier analysis fitted pseudo-critical points to an $L^{-3}$ shift,
treated $U_P\to2/3$ as first-order evidence, and interpreted a failed visual
collapse as excluding Ising universality. Those inferences are withdrawn.
With a coarse coupling grid, unequal statistics, and no propagated
autocorrelation errors, a poor collapse can arise from peak-location error,
scaling corrections, observable normalization, or insufficient equilibration.
It is not evidence for a different universality class.

The previously displayed double-peaked histograms were synthetic curves made
from means and variances, not sampled probability distributions. They have
been removed from the evidence chain. No latent heat or phase coexistence was
measured.

\subsection{What the Existing Data Support}

The current record supports only the following limited statements:
\begin{enumerate}
 \item the simulations resolve strong finite-size variation in the established
       critical region near $\beta\simeq0.76$;
 \item the local plaquette and its fluctuations provide useful equilibration
       and energy-like diagnostics;
 \item the data do not establish the order of the transition independently,
       nor do they demonstrate that a physical time-orientation observable
       becomes nonzero.
\end{enumerate}

Consequently, the numerical results should not be described as direct evidence
that a cosmological arrow of time emerges either sharply or continuously. Such
a claim requires a separately defined gauge-invariant physical observable and
its scaling behavior.

\subsection{Required Reanalysis}

A publishable reanalysis should retain time series, estimate integrated
autocorrelation times, use blocked or jackknife uncertainties, and analyze
several lattice sizes on a coupling grid fine enough for reweighting or
controlled interpolation. The finite-size ansatz should use 3D Ising exponents,
including corrections to scaling. Extended Wilson loops or an equivalent
nonlocal diagnostic must be analyzed separately from the plaquette. Until that
work is complete, the raw data are preserved as exploratory results and no
post-May T20d claim is promoted into the main manuscript.
